String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-30 2023-1-19/sections/algorithms.tex
add files
experiments

% \vspace{-3mm}
\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
% \footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
%$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ 
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
% $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})_{i \in [n], j \in [l_i]}]$ is true. 
%$\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}
% \vspace{-4mm}
\begin{proposition}\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
\begin{proof}
  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

%As mentioned in Section \ref{sec:overview}, the main idea of our approach is to model the counting operators symbolically by registers instead of unfolding them explicitly. Additionally, we use registers to represent string lengths. After that, the satisfiability of string constraints involves regex-counting, and string-length is reduced to the satisfiability of LIA, which off-the-shelf SMT solvers can solve. We also propose techniques to reduce automata sizes and utilize under approximations to enhance performance.
\vspace{-3mm}
\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of counting operators. After that, we utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 48,843 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. It turns out that we can solve almost all of them except 181.   
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

\vspace{-2mm}
\subsection{Solving $\cefadec$} \label{subsec:cefadec}

@@

\medskip
\noindent
\emph{Step 1.} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \hfill{$\blacksquare$}

@@

\vspace{-2mm}
\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}
\vspace{-2mm}

@@

Because the sizes of the LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ are proportional to the sizes (more precisely, the alphabet size, the number of states and transitions) of NFA $\cU_{\cB_1}, \cdots, \cU_{\cB_n}$, and the satisfiability of existential LIA formulas is expensive (NP-complete) to solve, we have to reduce the sizes of these NFA.

Actually, we find that 
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, i.e.,
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. Therefore, we can remove these transitions in step 2.

\medskip
\noindent
\emph{Step 2.} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.  \hfill{$\blacksquare$}

@@

\medskip
\noindent
\emph{Step 3.} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula.$\blacksquare$
%
\vspace{-1mm}
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
\vspace{-1mm}

%\[
%\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
%\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
%\]

>>>>

% \vspace{-2mm}
\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
% \footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
%$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ 
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
% $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})_{i \in [n], j \in [l_i]}]$ is true. 
%$\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}
% \vspace{-4mm}
\begin{proposition}\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
\begin{proof}
  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

%As mentioned in Section \ref{sec:overview}, the main idea of our approach is to model the counting operators symbolically by registers instead of unfolding them explicitly. Additionally, we use registers to represent string lengths. After that, the satisfiability of string constraints involves regex-counting, and string-length is reduced to the satisfiability of LIA, which off-the-shelf SMT solvers can solve. We also propose techniques to reduce automata sizes and utilize under approximations to enhance performance.
\vspace{-2mm}
\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}
\vspace{-1mm}

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of counting operators. After that, we utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*)^{\{2,\infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa+aaaa) \concat b^*)^{\{2, \infty\}}$. 
The unfoldings of the inner counting operators of non-register-representable regexes incur an exponential blow-up in the worst case. Nevertheless, those regexes occupy only 5\% of the 48,843 regexes that are collected from the practice (see Section~\ref{sec:bench}). Moreover, the unfoldings are partial in the sense that the outmost counting operators are not unfolded. 
It turns out that our approach can solve almost all the RECL constraints involving these 48,843 regexes, except 181 of them (See Table~\ref{tab:results_regcol}). 
%$(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

@@

\vspace{-2mm}
\subsection{Solving $\cefadec$} \label{subsec:cefadec}
\vspace{-1mm}

@@

\medskip
\noindent
\emph{Step 1.} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

@@

\vspace{-2mm}
\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}
\vspace{-1mm}

@@

Because the sizes of the LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ are proportional to the sizes (more precisely, the alphabet size, the number of states and transitions) of NFA $\cU_{\cB_1}, \cdots, \cU_{\cB_n}$, and the satisfiability of existential LIA formulas is NP-complete, it is vital to reduce the sizes of these NFA to improve the performance.

Since
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, it turns out that 
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. Therefore, we propose the following simplification technique for the NFA $\cU_{\cB_i}$'s.

\medskip
\noindent
\emph{Step 2.} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.   \qed

@@

\medskip
\noindent
\emph{Step 3.} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\vspace{-1mm}
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
\vspace{-1mm}
\qed
%\[
%\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
%\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
%\]
